# Animations - Two ways {#sec-animations}

::: {.callout-note appearance="simple"}
Today we focus on animating visualizations in R
:::

## Animations

Visualizations that include movement are a another way of creating salience. However, a bad animation doesn't add anything to the visualization and just requires more time to view the same information than a good visualizationn.

First, let's show a couple of examples using `gganimate` which works to extend the grammar of `ggplot2`. Then we will also show a couple of examples using `shiny` and its `sliderInput()` animation.

## Packages and libraries

Install `gganimate` and `gifski`. Apple computers may default to image-magick, but I can't test that.

```{r}
#| label: install gganimate and gifski
#| eval: false
#| echo: true

install.packages('gganimate')
install.packages('gifski') # Note that gifski may require other dependencies such as a Rust installation.  
```

Load libraries used for visualization today.

```{r}
#| label: load today's libraries
#| eval: true
#| echo: true

library(tidyverse) ## or dplyr and tidyr
library(sf)
library(gganimate)
library(gifski)
library(leaflet)
library(shiny)

```

## [`gganimate`](https://gganimate.com/)

### Contrived example - Keeling Curve at Mauna Loa

In [Lecture 1](https://radicalresearch.llc/EA078_Fall2024/information.html#what-is-the-baseline) we plotted the growing concentration of CO~2~ at Mauna Loa, the famous Keeling Curve.

Let's revisit that.

First, import the data from NOAA CMDL.

```{r}
#| label: ImportCO2
#| warning: false

#read raw data
co2 <- read_table('https://www.gml.noaa.gov/webdata/ccgg/trends/co2/co2_mm_mlo.txt',
                  skip = 57 ) 
#fix column headers
fieldNames <- c('year', 'month', 'decDate', 'meanCO2', 'trendedCO2', 'days', 'stdev', 'unc')
colnames(co2) <- fieldNames
# check dataset back rows
tail(co2)

```

Next, let's create a relatively simple visualization of the Keeling Curve using ggplot2. @fig-KeelingCurve shows the result.

```{r}
#| label: fig-KeelingCurve
#| fig-cap: Concentration (ppm) of CO2 measured at Mauna Loa
#| warning: false
#| echo: true

  ggplot(data = co2, aes(x = decDate, y = meanCO2)) +
  geom_point(color = 'black', size = 1) +
  theme_bw() +
  labs(x = 'Year', y = 'Concentration of CO2 (ppm)', 'Keeling Curve @ Mauna Loa')
```

### Add in the animation steps

In many environmental data sets, we will want to show changes over time. `gganimate` has a built-in function for time animations called `[transition_time()]`(https://gganimate.com/reference/transition_time.html).

Let's do the most super-basic animation and add that function to the Keeling Curve visualization. @fig-BasicAnimation shows the most basic animation when adding a time increment.

```{r}
#| label: fig-BasicAnimation
#| fig-cap: Animated Concentration (ppm) of CO2 measured at Mauna Loa
#| warning: false
#| echo: true

ggplot(data = co2, aes(x = decDate, y = meanCO2)) +
  geom_point(color = 'black', size = 1) +
  theme_bw() +
  labs(x = 'Year', y = 'Concentration of CO2 (ppm)', 'Keeling Curve @ Mauna Loa') +
  transition_time(year) # This step adds the by year animation.

```

Pretty cool, but we don't see the old data so it looks just like a migrating flock of points. If we want to show other points along the graph, we can use `shadow_mark()` to show other points along the graph. Arguments for past and future allow us to choose include either or both of those points.

`shadow_mark(past = TRUE, future = FALSE, ..., exclude_layer = NULL)`

@fig-ShadowAnimation shows the result, while adding in a color argument to shadow_mark to show the old data differently.

```{r}
#| label: fig-ShadowAnimation
#| fig-cap: Animated Concentration (ppm) of CO2 measured at Mauna Loa with shadowed past
#| warning: false
#| echo: true

ggplot(data = co2, aes(x = decDate, y = meanCO2)) +
  geom_point(color = 'black', size = 1) +
  theme_bw() +
  labs(x = 'Year', y = 'Concentration of CO2 (ppm)', 
       title = 'Year: {frame_time}') +
  transition_time(year) +
  shadow_mark(past = TRUE, color = 'gray') # this leaves behind the animated data already shown.  Feel free to choose a different color.

```

Pretty close, but that Year title is *horrible* and the ten significant figures past the decimal point is very aesthetically displeasing. I  must fix that using the `round()` function.

The interesting thing about that curly bracket notation is it can deal with variables and code directly. So let's modify that directly.

@fig-RoundYear shows the fixed title.

```{r}
#| label: fig-RoundYear
#| fig-cap: Animated Concentration (ppm) of CO2 measured at Mauna Loa with shadowed past and rounded Year in title.
#| warning: false

ggplot(data = co2, aes(x = decDate, y = meanCO2)) +
  geom_point(color = 'black', size = 1) +
  theme_bw() +
  labs(x = 'Year', y = 'Concentration of CO2 (ppm)', 
       title = 'Year: {round(frame_time, 0)}') + # the rounded value of the year is inserted here, with zero to make it an integer.
  transition_time(year) +
  shadow_mark(past = TRUE, color = 'grey') 

```

### Example 2: Animating a ggplot map

Import warehouse data for Riverside County only - let's limit the scope.

```{r}
#| label: Import warehouses data for Riverside County
#| warning: false
#| echo: true

WH.url <- 'https://raw.githubusercontent.com/RadicalResearchLLC/WarehouseMap/main/WarehouseCITY/geoJSON/comboFinal.geojson'
warehouses <- st_read(WH.url) |>  
  filter(county == 'Riverside County') |>  
  st_transform(crs = 4326)

head(warehouses)

```

Make a basic warehouse map near my house using ggplot and geom_sf. @fig-SimpleWHmap shows a basic map of warehouses in ggplot.

```{r}
#| label: fig-SimpleWHmap
#| fig-cap: A simple map of warehouses near Mike's house
#| warning: false
#| echo: true

ggplot(data = warehouses) +
  geom_sf() +
  coord_sf(xlim = c(-117.35, -117.1),
           ylim = c(33.8,33.95), crs = 4326) +
  theme_void() 
  
```

Let's animate it. We'll add a second step to control the animation speed and frames. First, we add `transition_time()` and `shadow_mark()` in a way identical to our CO~2~ figure.

Pass the ggplot code chunk into a variable. This variable is then run through an `animate()` function to control the frame rate and number of frames displayed.

@fig-TS_WHmap shows the time series animation.

```{r}
#| label: fig-TS_WHmap
#| fig-cap: An animated map of warehouses growing near Mike's house
#| warning: false
#| echo: true

data4map <- ggplot(data = warehouses) +
  geom_sf(fill = 'black') +
  coord_sf(xlim = c(-117.35, -117.1),
           ylim = c(33.8,33.95), crs = 4326) +
  theme_void() + 
  transition_time(year_built) + # new line to animate by year_built
  shadow_mark(past = TRUE, color = 'grey20', fill = 'grey') + # new line to leave built warehouses as grey blocks
  labs(title = 'Year: {round(frame_time, 0)}') # new title for figure showing the rounded year year

animate(data4map, nframes = 46, fps = 3, end_pause = 10) # this is a wrapper to control the speed of the animation.  
  
```

Excellent! We can also add an underlying map of jurisdictions or a tile layer to make it a bit prettier.

Let's use tigris places function to add some jurisdictions.

```{r}
#| warning: false
#| echo: true

library(tigris)

#read the data
Riverside_places <- places(state = 'CA', cb = T, #pull for California, lower detail level
                    year = 2022, 
                    progress_bar = FALSE) |> 
  filter(NAME %in% c('Moreno Valley', 'Mead Valley', 'March ARB', 'Perris', 'Riverside', 'Nuevo')) |> 
  st_transform(crs = 4326)
  
head(Riverside_places)

```

And this chapter from the [ggplot2 book](https://ggplot2-book.org/scales-colour) is much better than my lecture on color - please review.


```{r}
#| label: fig-PrettierMap
#| fig-cap: An animated map of warehouses growing near Mike's house with a background layer
#| warning: false
#| echo: true

data4map2 <- ggplot() +
  geom_sf(data = Riverside_places, aes(fill = NAME), alpha = 0.2) +
  geom_sf(data = warehouses, color = 'black', fill = '#653503',
          inherit.aes = FALSE) +
  coord_sf(xlim = c(-117.35, -117.1),
           ylim = c(33.7,33.95), crs = 4326) +
  theme_void() + 
  transition_time(year_built) +
  shadow_mark(past = TRUE, color = '#653503', fill = 'grey') +
  labs(title = 'Year: {round(frame_time, 0)}') +
  scale_fill_brewer(palette = 'Accent')

animate(data4map2, nframes = 47, fps = 3, end_pause = 10)
  
```
It could still be better adding roads and rail and other features and making the colors less ridiculous, but this is a good enough example for today.

## Animations in Shiny using `sliderInput()`

### Example 3 - Animate Old Faithful Histogram

Create a new shiny App called 'animate.R' as shown in [Section 19.3.2](https://mikem-radicalresearch.quarto.pub/environmental-data-visualization/shiny.html#fig-menuShiny).

This should create a new Shiny App using the Old Faithful Geyser data. We're going to animate this using the following argument within the `sliderInput()` function.

`animate = TRUE`.

The code chunk within the app should look like this, starting at line 21 on my machine:

```{r}
#| label: animate shiny app code chunk 1
#| eval: false
#| echo: true

            sliderInput("bins",
                        "Number of bins:",
                        min = 1,
                        max = 50,
                        value = 30,
                        # NEW ARGUMENT HERE!
                        animate = TRUE)
```

If we run the app by pressing the Run App button, a shiny App should pop-up. I'll show that within the Shiny App. A blue **play** button will appear on the bottom-right of the slider. Pressing the play button advances through the slider increments and updates the histogram.

Easy!

There are additional options for controlling the interval rate and whether it loops.

We can modify the code to show that.

```{r}
#| label: animate shiny app code chunk 2
#| eval: false
#| echo: true

            sliderInput("bins",
                        "Number of bins:",
                        min = 1,
                        max = 50,
                        value = 30,
                        # NEW ARGUMENT HERE!
                        animate = animationOptions(
                          loop = TRUE, interval = 300)
                        )
```
