# Roads {#sec-roads}

::: {.callout-note appearance="simple"}
Today is going to be an showing how to get some roads data.  
:::
  
## Load libraries

```{r}
library(tigris)
library(sf)
library(leaflet)
library(dplyr)
```
```{r}
knitr::opts_chunk$set(message = NA)
```

### `Tigris` function for scraping road data.

The aptly named `roads()` function will pull road linestrings.  

```{r}
#| messages: false
#| warnings: false
Riv_roads <- roads(state = 'CA', county = 'Riverside', year = 2023, progress_bar = FALSE) |> 
  st_transform(crs = 4326)

head(Riv_roads)

```
There are 58,448 road segments from 2023.  That's a lot of lines for a map and takes too long to load.  Feel free to make a map with it if you want to hear your computer hum.  

For the purpose of the road widening project, we're mostly focusing on big roads getting widened.  The documentation for `roads()` states that the RTTYP category provides six types.  

I'm going to `filter` just the Interstate (I) and State (S) roads for this version.  

```{r}

Riv_major_roads <- Riv_roads |> 
  filter(RTTYP %in% c('I', 'S'))

```

There are only 81 major roads segments that meet that criteria.  

```{r}
#| label: fig-Roads
#| echo: true
#| warning: false
#| fig-cap: Riverside roads on a leaflet map

  leaflet() |> 
  addTiles() |> 
  addPolylines(data = Riv_major_roads,
              color = 'black',
              weight = 3,
              dashArray = '2,6', #new feature - makes lines dashed - line lenght, space length
              label = ~FULLNAME)
```

Now let's add Riverside County warehouses for a quick look at the proximity of those facilities to current major roadways.  We did this in [lesson 10.3.2](https://radicalresearch.llc/EA078_Fall2024/EJPolygons.html#manually-identify-the-polygon-vertices.) 

```{r}


WH.url <- 'https://raw.githubusercontent.com/RadicalResearchLLC/WarehouseMap/main/WarehouseCITY/geoJSON/comboFinal.geojson'
warehouses <- st_read(WH.url) |>  
  st_transform(crs = 4326) |> 
  filter(county == 'Riverside County')

```

@fig-roadWH shows the major roads and warehouse land-use.

```{r}
#| label: fig-roadWH
#| echo: true
#| warning: false
#| fig-cap: Riverside roads and warehouses on a leaflet map

palWH <- colorFactor(palette = c('orange', 'gold', 'brown'),
                     domain = warehouses$category)
palRoadType <- colorFactor(palette = c('darkblue', 'purple'),
                           domain = Riv_major_roads$RTTYP)

leaflet() |> 
  addProviderTiles(provider = providers$CartoDB.Positron) |> ## We modified this line to change to satellite view  
  addPolygons(data = warehouses,
              color = ~palWH(category),
              weight = 1,
              fillOpacity = 0.7) |> 
  addLegend(data = warehouses,
            pal = palWH,
            values = ~category) |> 
  addPolylines(data = Riv_major_roads,
            color = ~palRoadType(RTTYP),
            weight = 3,
            dashArray = '2,6',
            label = ~FULLNAME) |> 
  setView(zoom = 10, lat = 33.8, lng = -117.15) 
```

This is a good start, but there's some missing pieces that we can show a few examples for road widening.

Let's `filter()` down a data table for a couple of key roads that are approved or under review for road widening.

```{r}

roads2widen <- Riv_roads |> 
  filter(FULLNAME %in% c('Cajalco Rd', 'Cajalco Expy', 'Ramona Expy', 'Gilman Springs Rd'))

```

This selects just a few individual county level roads.  Let's add those to a new map.

@fig-widening shows the result.

```{r}
#| label: fig-widening
#| echo: true
#| warning: false
#| fig-cap: Riverside roads and widening roads with warehouses on a leaflet map

palWH <- colorFactor(palette = c('orange', 'gold', 'brown'),
                     domain = warehouses$category)
palRoadType <- colorFactor(palette = c('darkblue', 'purple'),
                           domain = Riv_major_roads$RTTYP)

leaflet() |> 
  addProviderTiles(provider = providers$CartoDB.Positron) |> ## We modified this line to change to satellite view  
  addPolygons(data = warehouses,
              color = ~palWH(category),
              weight = 1,
              fillOpacity = 0.7) |> 
  addLegend(data = warehouses,
            pal = palWH,
            values = ~category,
            position = 'bottomleft') |> 
  addPolylines(data = Riv_major_roads,
            color = ~palRoadType(RTTYP),
            weight = 3,
            dashArray = '2,6',
            label = ~FULLNAME) |> 
  setView(zoom = 10, lat = 33.8, lng = -117.15) |> 
  addPolylines(data = roads2widen,
               color = 'red',
               weight = 4,
              dashArray = '2,6',
              label = ~FULLNAME) |> 
  addLegend(title = 'Roads',
            colors = c('darkblue', 'purple', 'red'),
            labels = c('Interstate', 'State Route', 'Road widening'),
            position = 'bottomleft')
```
